---
title: "Deaths due to Flu/Pneumonia and COVID in US"
author: "Tomasz Przechlewski"
date: "2020-12-22"
output:
  html_document:
    includes:
      before_body: up_hdr.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data

Data is from National Center for Health Statistics (NCHS) Mortality Surveillance
available in CSV format at https://www.cdc.gov/flu/weekly/. 
(URL `https://www.cdc.gov/flu/weekly/weeklyarchives2020-2021/data/NCHSData??csv`where `??` is a week
number.)

There are 7 variables in CSV file:
year; week;
pip (Percent of Deaths Due to Pneumonia and Influenza (P&I));
pic (Percent of Deaths Due to Pneumonia, Influenza or COVID-19 (PIC));
expected (Expected);
threshold (Threshold);
total (All Deaths);
pneumonia (Pneumonia Deaths);
influenza (Influenza Deaths),
covid19 (COVID-19 Deaths),
PIC (Pneumonia, Influenza or COVID-19 Deaths (PIC))

```{r, echo=FALSE, warning=F, message=F}
library("dplyr")
library("ggplot2")
library("ggpubr")
library("scales")
##
mainBreaks <- "6 months"
surl <- "https://www.cdc.gov/flu/weekly/#S2"
spanV <- 0.5
```

The R code for reading CSV file is standard:

```{r }
d <- read.csv("us.pneumonia.csv", sep = ';',  header=T, na.string="NA")
# wdate column is a string "yyyy/ww"
d$wdate <- sprintf ("%.4i/%.02i", d$year, d$week)
first.obs <- first(d$wdate)
last.obs <- last(d$wdate)
```

This first week is `r first.obs` and the last is `r last.obs`.


Using `year` and `week` we compute date as defined as first day of each week:
(`%W` is week number of the year):

```{r}
## Note if insufficient arguments provided to strptime, current date
## will be returned (ie %Y/%W is insufficent)
d$date <- strptime(sprintf("%s-%s-01", d$year, d$week), format="%Y-%W-%w")
```

## Charts

Weekly number of deaths due to Influenza:

```{r, echo=F, warning=F}
pc1 <- ggplot(d, aes(x= as.Date(date), y=influenza )) +
 geom_line(aes(), group=1, size=.4, color="steelblue") +
 ##geom_smooth(method="loess", se=F, span=spanV) +
 xlab(label="year/month") +
 ylab(label="deaths/week") +
 scale_x_date(labels = date_format("%y-%m"), breaks = mainBreaks)+
 theme(plot.subtitle=element_text(size=8, hjust=0, color="black")) +
 ggtitle(sprintf("Deaths due to influenza in US (last week: %s)", last.obs), 
         subtitle=sprintf("Source: %s", surl)) 
pc1
```

Weekly number of deaths due to Pneumonia, Influenza or COVID:

```{r, echo=F, warning=F}
pd1 <- ggplot(d, aes(x = as.Date(date), y=PIC ) ) +
 geom_line(aes(), group=1, size=.4, color="steelblue") +
 ##geom_smooth(method="loess", se=F, span=spanV) +
 xlab(label="year/month") +
 ylab(label="deaths/week") +
 scale_x_date(labels = date_format("%y-%m"), breaks = mainBreaks)+
 theme(plot.subtitle=element_text(size=8, hjust=0, color="black")) +
 ggtitle(sprintf("Deaths due to pneumonia/influenza/covid in US (last week: %s)", last.obs), subtitle=sprintf("Source: %s", surl)) 
##
##ggsave(pc1, file="us_influenza.png", width=10)
##ggsave(pd1, file="us.pic.png", width=10)
pd1
```

The same chart as above, but this time dot-plot with trend and weeks of high level flu 
highlighted.
We define flu season as first 12 weeks of each year. To highlight these periods we define
the following dataframe:

```{r, echo=T, warning=F}

bars <- data.frame(
   x1=c("2021-01-01", "2020-01-01", "2019-01-01", "2018-01-01", "2017-01-01", "2016-01-01", "2015-01-01", "2014-01-01"), 
   x2=c("2021-03-01", "2020-03-01", "2019-03-01", "2018-03-01", "2017-03-01", "2016-03-01", "2015-03-01", "2014-03-01"), 
   y1=c(0, 0, 0, 0, 0, 0, 0, 0), 
   y2=c(Inf,Inf, Inf, Inf, Inf, Inf, Inf, Inf)
)
```

Dots in flu season are drown with different color. To achieve this we define
column `fs` as follows:

```{r}
d <- d %>% mutate( fs=case_when(week < 12 ~ "1", TRUE ~ "0"))
```

Next we add `aes(color=as.factor(fs))` (consult the code in source Rmd file) and `set_color_manual`
to add legend. The legend is put under title with the declaration
`theme_update(legend.title=element_text(family="sans", size=8), legend.position="top" )`.

Crucial is `inherit.aes = FALSE` in `geom_rect` otherwise R complains about errors.

```{r, echo=F, warning=F, message=F}
#theme_update(legend.title=element_text(family="sans", size=8), legend.position="top" )

pe1 <- ggplot(d, aes(x = as.Date(date), y=PIC)) +
        geom_point(aes(color=as.factor(fs)), size=.4) +
        geom_smooth(method="loess", se=F, span=spanV ) +
        xlab(label="year/month") +
        ylab(label="deaths/week") +
        scale_color_manual("Tygodnie/Weeks:", values = c( "1" = "#F8766D", "0" = "#00BFC4" ), 
                           labels=c("1"="01--12 (sezon grypowy/flu season)", "0" = "13--52/3") ) +
        scale_x_date(labels = date_format("%y-%m"), breaks = mainBreaks)+
        theme(plot.subtitle=element_text(size=8, hjust=0, color="black")) +
        ##theme_nikw() +
        ## inherit.aes =F bez tego nie działa
        geom_rect(data=bars, inherit.aes = FALSE,
                  mapping=aes(xmin=as.Date(x1), 
                              xmax=as.Date(x2), ymin=y1, ymax=y2), 
                  fill = "#F8766D", alpha = 0.25) +
        ggtitle(sprintf("Zgony z powodu zapalenia płuc/grypy/covid19  (USA; ostatni tydzień: %s)", last.obs), 
          subtitle=sprintf("Deaths due to pneumonia/influenza/covid19 (US; last week  %s)\n Źródło/Source: %s", last.obs, surl)) 

pe1
##ggsave(pe1, file="us.pic.png", width=8)
```

In case of errors noticed or with suggestions for improvement please contact 
tprzechlewski[a_t]gmail.com

